Skill estimation using TrueSkill Through Time: \\ the \proglang{Julia}, \proglang{Python} and \proglang{R} packages (Draft)

### 2021-05-10 doc/article.tex
Update documentation

\usepackage{wrapfig}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{amsmath} %para escribir funci\'on partida , matrices
\usepackage{amsthm} %para numerar definciones y teoremas
\usepackage{amsfonts} % \mathbb{N} -> conjunto de los n\'umeros naturales
\usepackage{bm} % \bm{\alpha} bold greek symbol
\usepackage[makeroom]{cancel} % \cancel{} \bcancel{} etc
\usepackage{mdframed}
\usepackage{algorithm}
\usepackage{comment}
%\usepackage{quoting}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{csvsimple}
\usepackage{listings}
\renewcommand{\lstlistingname}{Code}% Listing -> Algorithm
\usepackage{todonotes}
\usepackage[binary-units]{siunitx}

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Skill estimation using TrueSkill Through Time: \\ the \proglang{Julia}, \proglang{Python} and \proglang{R} packages (Draft)}
\Plaintitle{Skill estimation using TrueSkill Through Time:  \\ the Julia, Python and R packages (Draft)}
\Shorttitle{TrueSkill Through Time: the \proglang{Julia}, \proglang{Python} and \proglang{R} packages (Draft)}

@@

  Esteban Mocskos\\
  Departamento de Computaci\'on\\
  Facultad de Ciencias Exactas y Naturales\\
  Universidad de Buenos Aires\\
  Buenos Aires, Argentina\\
  \emph{and}\\
  Departamento de Computaci\'on\\
  Facultad de Ciencias Exactas y Naturales\\
  Universidad de Buenos Aires\\
  Buenos Aires, Argentina\\
}

%% - \Abstract{} almost as usual
\Abstract{
\en{Humans develop complex skills through time.}
 \es{Los humanos desarrollamos habilidades complejas a trav\'es del tiempo.}
 %
 \en{Estimating them is essential in education, labor and sports systems.}
 \es{Estimarlas es esencial en los sistemas educativos, laborales y deportivos.}
 %
 \en{However, it is a sensitive issue when they are intended to be used to make decisions that may impact individuals.}
 \es{Sin embargo, es un tema sensible cuando se las pretende usar para tomar decisiones que puede impactar a los individuos.}
 %
 \en{None of the widely used skill estimators, such as TrueSkill, Glicko and Item-Response Theory, correctly models the temporal aspect, which prevents having both good initial estimates and comparability between estimates separated in time and space.}
 \es{Ninguno de los estimadores habilidad ampliamente utilizados, como TrueSkill, Glicko e Item-Response Theory, modela correctamente el aspecto temporal, lo que les impide tener tanto buenas estimaciones iniciales como comparabilidad entre estimaciones separadas en el tiempo y en el espacio.}
 %
 \en{TrueSkill Through corrects those biases by modeling the entire history of activities using a single Bayesian network.}
 \es{TrueSkill Through Time corrige esos sesgos al modelar toda la historia de actividades mediante única red bayesiana.}
 %
 \en{The use of an efficient algorithm, that requires only a few linear iterations over the data, allows scaling to millions of observations in few seconds.}
 \es{El uso de un algoritmo eficiente, que requiere s\'olo unas pocas iteraciones lineales sobre los datos, permite escalar a millones de observaciones en pocos segundos.}
 %
 \en{In this paper we offer the first packages for \proglang{Julia}, \proglang{Python}, and \proglang{R}, together with its scientific documentation, supporting the use of reliable learning curve estimators.}
 \es{En este artículo ofrecemos los primeros paquetes para \proglang{Julia}, \proglang{Python} y \proglang{R}, acompañada con su documentaci\'on cient\'ifica, apoyando el uso de estimadores de curvas de aprendizaje confiables.}
}
\Keywords{
\en{Learning, skill, gaming, education, sports, \proglang{Julia}, \proglang{Python},  \proglang{R}}
\es{Aprendizaje, habilidad, videojuego, educación, deporte, \proglang{Julia}, \proglang{Python},  \proglang{R}}
}
\Plainkeywords{}

@@

\en{Humans develop complex skills because of an special integration of biological, cognitive and social processes~\citep{Koster2020}.}
\es{Los humanos desarrollamos habilidades complejas gracias a una integraci\'on especial de los procesos biol\'ogicos, cognitivos y sociales~\citep{Koster2020}.}
%
\en{Our exceptional cognitive ability to imitate, combined with long periods of juvenile dependency and postreproductive life span, allows humans to learn things from others and transmit innovations through generations~\citep{Richerson2020}.}
\es{Nuestra extraordinaria capacidad para imitar, combinada con los largos per\'iodos de aprendizaje juvenil y vida posreproductiva, permite a los humanos aprender de los dem\'as y transmitir las innovaci\'on a trav\'es de la generaciones~\citep{Richerson2020}.}
%
\en{As a population-based process, human adaptation is also affected by demographic characteristics, such as the size and structure of populations~\citep{Derex2020}.}
\es{Al ser un proceso poblacional, la adaptaci\'on humana tambi\'en se ve afectada por caracter\'isticas demogr\'aficas, como el tama\~no y estructura de las poblaciones~\citep{Derex2020}.}

\en{Knowing how individual skills change over time is essential in the educational system and in the labor market.}
\es{Conocer c\'omo cambian las habilidades individuales a lo largo del tiempo es esencial en el sistema educativo y laboral.}
%
\en{Since skills are hidden variables, the best we can do is to estimate it based on its direct observable consequences: the outcome of problem solving and competitions.}
\es{Dado que son variables ocultas, lo mejor que podemos hacer es estimarlas a partir de sus consecuencias observables directas: el producto de resoluci\'on de problemas y competencias.}
%
\en{However, estimating learning curves is a sensitive issue, especially when they are intended to be used to make decisions that may impact individuals.}
\es{Sin embargo la estimación de curvas de aprendiaje es tema sensible, especialmente cuando se las pretende utilizar para tomar decisiones que pueden impactar a las personas.}
%
\en{Considering only the frequency of positive results as an indicator of the individuals' ability could lead to wrong approximations, mainly because the outcome also depends on the difficulty of the challenge.}
\es{Considerar s\'olo la frecuencia de resultados positivos como indicador de la habilidad de los individuos puede conducir a aproximaciones erroneas, fundamentalmente porque su valor depende tambi\'en de la dificultad de los desaf\'ios.}
%
\en{For this reason, all widely used skill estimators are based on pairwise comparisons.}
\es{Por esta raz\'on, todos los estimadores de habilidad ampliamente usados se basan en comparaciones por pares.}
%
\en{Since the first generative models, proposed almost a century ago by~\cite{Thurstone1927} and~\cite{Zermelo1929}, it is assumed that the probability of the observed result $r$ depends on the performance $p$ of an agent $i$ and their opponent $j$, expressed as $P(\, r \,|\, p_i, \, p_j \,)$.}
\es{Desde los primeros modelos generativos, propuestos hace casi un siglo por~\cite{Thurstone1927} y~\cite{Zermelo1929}, se supone que la probabilidad de un resultado observado $r$ depende del rendimiento $p$ del agente $i$ y de su oponente $j$, expresada como $P(\, r \,|\, p_i, \, p_j \,)$.}
%
\en{The field continued to progress with the work of~\cite{Bradley1952} and~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}, leading to a breakthrough that took place when~\cite{Elo2008} developed a methodology for the US Chess Federation (USCF), which is still used by the International Chess Federation (FIDE).}
\es{El campo sigui\'o progresando con los trabajos de \cite{Bradley1952} y~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}, que condujeron al gran avance que tuvo lugar cuando~\cite{Elo2008} desarrolló una metodología para la Federaci\'on de Ajedrez de los Estados Unidos (USCF), adoptada hasta el d\'ia de hoy por la Federaci\'on Internacional de Ajedrez (FIDE).}

@@

\en{All currently used skill estimators share some variant of the probabilistic model proposed by Elo~\citep{Glickman1999,Herbrich2007,VanDerLinden2016,Fox2010}.}
\es{Todos los estimadores de habilidad ampliamente utilizados en la actualidad comparten alguna variante del modelo probabil\'istico propuesto por Elo~\citep{Glickman1999, Herbrich2007, VanDerLinden2016, Fox2010}.}
%
\begin{figure}[ht!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r = (d > 0)$};

@@

    \edge {d} {r};
    \edge {p1,p2} {d};
    \edge {s1} {p1};
    \edge {s2} {p2};
    %\node[invisible, right=of p2, xshift=4.35cm] (s-dist) {};
}
     \caption{
     \en{Generative model in which skills cause the observable results mediated by the difference of hidden performances, $d =p_i - p_j$, both random variables around their unknown true skill, $p \sim \N(s,\beta^2)$.}
    \es{Modelo generativo en el que las habilidades causan los resultados observables a trav\'es de la diferencia de rendimientos ocultos, $d=p_i-p_j$, ambas variables aleatorias centradas en la verdadera habilidad, $p \sim \N(s,\beta^2)$.}
    %
    \en{The one with the highest performance wins, $r = (d > 0)$.}
    \es{Quien haya obtenido mayor rendimiento gana, $r = (d > 0)$.}
    %
    \en{Observable variables are painted gray, hidden in white, and constants are shown as black dots.}
    \es{Las variables observables se pintan de gris, las ocultas en blanco, y las constantes se muestran como puntos negros.}
    }
    \label{fig:generative_model}
\end{figure}
%
\en{Figure~\ref{fig:generative_model} provides a causal model in which skills generates the observable result.}
\es{La figura~\ref{fig:generative_model} ofrece un modelo causal en el que las habilidades generan el resultado observable.}
%
\en{The agents exhibit different performances at each event, varying around their true skill, $\N(p\,|\,s,\beta^2)$.}
\es{Los agentes exhiben distintos desempe\~nos en cada evento, que var\'ian alrededor de su verdadera habilidad, $\N(p\,|\,s,\beta^2)$.}
%
\en{The model assumes that the agent with the highest performance wins, $r = (p_i > p_j)$. In other words, whoever obtains a difference of performance greater than 0 wins, $r = (p_i - p_j > 0)$.}
\es{El modelo supone que gana el agente con mayor rendimiento, $r = (p_i > p_j)$. En otras palabras, gana quien obtenga una diferencia de desempeño mayor a 0, $r = (p_i - p_j > 0)$.}
%
\en{The parameter $\beta^2$ is the same for all the agents and acts as the scale of the estimates: skills separated by a $\beta$ keep always the same probability of winning.}
\es{El parámetro $\beta^2$, al ser el mismo para todos los agentes, act\'ua como la escala de las estimaciones: habilidades separada por un $\beta$ mantienen siempre la misma probabilidad de ganar.}

@@

\en{Using the graphical representation proposed in Figure~\ref{fig:generative_model}, it is possible to derive the prediction of the result given the previous estimates, $P(\,r\,|\,s_{i_\text{old}},s_{j_\text{old}}\,)$.}
\es{A partir de la representaci\'on gr\'afica propuesta en la figura \ref{fig:generative_model}, se puede derivar la predicci\'on del resultado dadas las estimaciones previas, $P(\,r\,|\,s_{i_\text{old}},s_{j_\text{old}}\,)$.}
\en{Elo's methodological solution is simple and smart: updates both previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, based on the surprise, i.e. the complement of the prediction of the observed result.}
\es{La soluci\'on metodol\'ogica de Elo es simple y astuta: actualizar las estimaciones previas, $s_{i_\text{old}}$ y $s_{j_\text{old}}$, en base a la sorpresa, i.e. el complemento de la predicci\'on del resultado observado.}

@@

\en{where the probability arises from instantiating the previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, in the generative model (details in section~\ref{sec:2vs2}).}
\es{Donde la probabilidad surge de instanciar las estimaciones previas en el modelo generativo (detalles en la secci\'on~\ref{sec:2vs2}).}
\en{The idea is that the magnitude of the surprise $\Delta$ is related to the accuracy of the previous estimates and, therefore, can be used to update them.}
\es{La idea es que la magnitud de la sorpresa $\Delta$ est\'a relacionada con cuan buenas son las estimaciones previas, y por lo tanto puede usarse para actualizarlas.}
\en{Unexpected results would indicate that current estimates should be updated to a greater extent than if they had occurred as expected.}
\es{Resultados inesperados indicar\'ian que las estimaciones actuales no son del todo correctas y deber\'ian actualizarse en mayor medida que si hubieran ocurrido como se esperaba.}

@@

\en{where the surprise $\Delta$ acts as a correction factor for both previous estimates.}
\es{Donde la sorpresa $\Delta$ act\'ua como factor de correcci\'on para ambas estimaciones previas.}
\en{This procedure can recover the relative scale of the agents, starting from arbitrary initial values.}
\es{Esta soluci\'on puede recuperar la escala relativa de los agentes, partiendo de valores iniciales arbitrarios.}
\en{However, it does have a major weakness.}
\es{Sin embargo, tiene algunas debilidades importantes.}
\en{The update rule (Eq.~\ref{eq:elo_update}) is symmetric: what one agent wins is lost by the other.}
\es{La regla de actualizaci\'on (Eq.~\ref{eq:elo_update}) es sim\'etrica, as\'i que lo que gana un agente el otro lo pierde.}
\en{Because new agents start with arbitrary skills (the same initial value is used for all individuals),  they tend to generate high surprise and therefore can abruptly modify their opponents' estimates even though they have already converged.}
\es{Debido a que a los agentes nuevos comienzan con estimaciones arbitrarias (el mismo valor inicial para cualquier individuo), ellas tienden a generan alta sorpresa y por lo tanto pueden modificar bruscamente las estimaciones de sus oponentes a pesar de que ya hubieran convergido.}
\en{This weakness emerges because the uncertainties of the agent's estimates are not considered.}
\es{Esta debilidad ocurre por no tener en cuenta la incertidumbre sobre las estimaciones de los agentes.}
\en{An ad-hoc solution was proposed to solve this problem: reducing the impact of the surprise based on the number of times the agent has participated previously.}
\es{Una soluci\'on ad-hoc fue propuesta para resolver este problema: reducir el impacto de la sorpresa en funci\'on de la cantidad de veces que el agente ha participado previamente.}
% \en{This is the role played by the K-factor used by FIDE,  $\Delta_i = \Delta \cdot K_i$.}
% \es{Ese es rol que desempe\~na el K-factor usado por la FIDE, $\Delta_i = \Delta \cdot K_i$.}

@@

\en{A better alternative is to distribute our belief among all possible skill hypotheses, instead of selecting a single value as an estimate.}
\es{Una alternativa superadora consiste en distribuir nuestra creencia entre todas las posibles hipótesis de habilidad, en vez de seleccionar un \'unico valor como estimaci\'on.}
\en{This approach known as \emph{Bayesian inference} has proven successful in practice~\citep{Bishop2006} and ensures consistent reasoning in uncertainty contexts~\citep{Jaynes2003,VanHorn2003}.}
\es{Este enfoque conocido como inferencia bayesiana ha demostrado ser exitoso en la pr\'actica~\citep{Bishop2006} y garantiza un razonamiento consistente en contextos de incertidumbre~\citep{Jaynes2003,VanHorn2003}.}
\en{Any inference of this type, independently of its complexity, can be solved by the rules of probability: the~\ref{eq:sum_rule} and the~\ref{eq:product_rule}.}
\es{Toda inferencia de este tipo, no importa cuan compleja sea, puede ser resuelta mediante las reglas de la probabilidad: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}.}
\en{The \ref{eq:sum_rule} states that any marginal distribution can be obtained by integrating, or summing up, the joint distribution:}
\es{La \ref{eq:sum_rule} afirma que cualquier distribuci\'on marginal se puede obtener integrando o sumando la distribuci\'on conjunta.}

@@

\en{where $ P(\cdot)$ and $p(\cdot)$ represent discrete and continuous probabilities distributions respectively.}
\es{Donde $P(\cdot)$ y $p(\cdot)$ representan distribuciones de probabilidad discretas y continuas respectivamente.}
\en{Additionally, the \ref{eq:product_rule} states that any joint distribution can be expressed as the product of one-dimensional conditional distributions.}
\es{Adem\'as, la \ref{eq:product_rule} se\~nala que cualquier distribuci\'on conjunta puede ser expresada como el producto de distribuciones condicionales uni-dimensionles.}

@@

% \en{The inferential use of Bayes' theorem plays a central role in modern statistical learning techniques.}
% \es{El uso inferencial del teorema de bayes juega un rol central en las t\'ecnicas modernas de aprendizaje estad\'isitico.}
\en{The \ref{eq:bayes_theorem} optimally update our beliefs about the hypotheses,  maximizing uncertainty after incorporating the information provided by the model and the data.}
\es{El \ref{eq:bayes_theorem} actualiza de forma \'optima las creencia sobre las hip\'otesis, maximizando la incertidumbre luego de haber incorporar la información provista por el modelo y los datos.}

@@

\en{where the only free variable is the skill hypothesis of agent $i$.}
\es{Donde la \'unica variable libre es la hip\'otesis de habilidad del agente $i$.}
\en{The prior quantifies the uncertainty about the skill before seeing the result, and the posterior quantifies the uncertainty after seeing the result.}

>>>>

\usepackage{wrapfig}
\usepackage{caption}
\usepackage[labelformat=simple]{subcaption}
\renewcommand\thesubfigure{(\alph{subfigure})}

\usepackage{amsmath} %para escribir funci\'on partida , matrices
\usepackage{amsthm} %para numerar definciones y teoremas
\usepackage{amsfonts} % \mathbb{N} -> conjunto de los n\'umeros naturales
\usepackage{bm} % \bm{\alpha} bold greek symbol
\usepackage[makeroom]{cancel} % \cancel{} \bcancel{} etc
\usepackage{mdframed}
\usepackage{algorithm}
\usepackage{comment}
%\usepackage{quoting}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{csvsimple}
\usepackage{listings}
\renewcommand{\lstlistingname}{Code}% Listing -> Algorithm
\usepackage{todonotes}
\usepackage[binary-units]{siunitx}

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Skill estimation using TrueSkill Through Time: \\ the \proglang{Julia}, \proglang{Python} and \proglang{R} packages (Draft)}
\Plaintitle{Skill estimation using TrueSkill Through Time: the Julia, Python and R packages (Draft)}
\Shorttitle{TrueSkill Through Time: the \proglang{Julia}, \proglang{Python} and \proglang{R} packages (Draft)}

@@

  Esteban Mocskos\\
  Departamento de Computaci\'on\\
  Facultad de Ciencias Exactas y Naturales\\
  Universidad de Buenos Aires\\
  Buenos Aires, Argentina\\
  \emph{and}\\
  Centro de Simulaci\'on Computacional p/Aplic Tecnol\'ogicas, CSC-CONICET\\
  Buenos Aires, Argentina\\
  E-mail: \texttt{emocskos@dc.uba.ar}\\
}

%% - \Abstract{} almost as usual
\Abstract{
\en{Humans develop complex skills through time. }%
\es{Los humanos desarrollamos habilidades complejas a través del tiempo. }%
%
\en{Knowing how individual abilities change is essential in a wide range of activities. }%
\es{Saber cómo cambian las capacidades individuales es esencial en diversas actividades. }%
%
\en{All widely used skill estimators, such as TrueSkill, Glicko and Item-Response Theory, are based on the outcome of problem-solving and competitions. }%
\es{Todos los estimadores de habilidades ampliamente utilizados, como TrueSkill, Glicko e Item-Response Theory, se basan en el producto de resolucion de problemas y competencias. }%
%
\en{However, they do not properly model the temporal aspect, which prevents them from achieving both good initial estimates and comparability between distant estimates. }%
\es{Sin embargo, ellos no modelan apropiadamente el aspecto temporal, lo que les impide alcanzar tanto buenas estimaciones iniciales como comparabilidad entre estimaciones distantes. }%
%
\en{TrueSkill Through Time correct those issues by modeling the entire history of activities using a single Bayesian network. }%
\es{TrueSkill Through Time corrige esos sesgos modelando toda la historia de actividades mediante una única red bayesiana. }%
%
\en{In this paper we offer the first packages for \proglang{Julia}, \proglang{Python}, and \proglang{R}, together with the scientific documentation, supporting the adoption of a reliable learning estimator. }%
\es{En este artículo ofrecemos los primeros paquetes para \proglang{Julia}, \proglang{Python} y \proglang{R}, acompañada con su documentaci\'on cient\'ifica, apoyando el uso de un estimador de aprendizaje confiable. }%
%
\en{We illustrate how to use them in typical cases, and show the actual learning curves of historical players of the Association of Tennis Professionals (ATP). }%
\es{Ilustramos cómo usarlos en casos típicos, y mostramos las curvas de aprendizaje reales de los jugadores históricos de la Asociación de Tenistas Profesionales (ATP).  }%
%
\en{This algorithm requires only a few iterations over the data, allowing it to scale to millions of observations in a few seconds. }%
\es{Este algoritmo requiere sólo unas pocas iteraciones sobre los datos, permitiendo escalar a millones de observaciones en pocos segundos. }%
}
\Keywords{%
\en{Learning, skill, gaming, education, sports, \proglang{Julia}, \proglang{Python},  \proglang{R}}%
\es{Aprendizaje, habilidad, videojuego, educación, deporte, \proglang{Julia}, \proglang{Python},  \proglang{R}}%
}
\Plainkeywords{%
\en{Learning, skill, gaming, education, sports, Julia, Python,  R}%
\es{Aprendizaje, habilidad, videojuego, educación, deporte, Julia, Python, R}%
}

@@

\en{Humans develop complex skills because of an special integration of biological, cognitive and social processes~\citep{Koster2020}. }%
\es{Los humanos desarrollamos habilidades complejas gracias a una integraci\'on especial de los procesos biol\'ogicos, cognitivos y sociales~\citep{Koster2020}. }%
%
\en{Our exceptional cognitive ability to imitate, combined with long periods of juvenile dependency and postreproductive life span, allows humans to learn things from others and transmit innovations through generations~\citep{Richerson2020}. }%
\es{Nuestra extraordinaria capacidad para imitar, combinada con los largos per\'iodos de aprendizaje juvenil y vida posreproductiva, permite a los humanos aprender de los dem\'as y transmitir las innovaci\'on a trav\'es de la generaciones~\citep{Richerson2020}. }%
%
\en{As a population-based process, human adaptation is also affected by demographic characteristics, such as the size and structure of populations~\citep{Derex2020}. }%
\es{Al ser un proceso poblacional, la adaptaci\'on humana tambi\'en se ve afectada por caracter\'isticas demogr\'aficas, como el tama\~no y estructura de las poblaciones~\citep{Derex2020}. }%

%
%
\en{Knowing how individual skills change over time is essential in the educational system and in the labor market. }%
\es{Conocer c\'omo cambian las habilidades individuales a lo largo del tiempo es esencial en el sistema educativo y laboral. }%
%
\en{Since skills are hidden variables, the best we can do is estimating them based on its direct observable consequences: the outcome of problem-solving and competitions. }%
\es{Dado que son variables ocultas, lo mejor que podemos hacer es estimarlas a partir de sus consecuencias observables directas: el producto de la resoluci\'on de problemas y competencias. }%
%
\en{However, estimating learning curves is a sensitive issue, especially when they are intended to be used to make decisions that may impact individuals. }%
\es{Sin embargo la estimación de curvas de aprendiaje es tema sensible, especialmente cuando se las pretende utilizar para tomar decisiones que pueden impactar a las personas. }%
%
\en{Considering only the frequency of positive results as an indicator of the individuals' ability could lead to wrong approximations, mainly because the outcome also depends on the difficulty of the challenge. }%
\es{Considerar s\'olo la frecuencia de resultados positivos como indicador de la habilidad de los individuos puede conducir a aproximaciones erroneas, fundamentalmente porque su valor depende tambi\'en de la dificultad de los desaf\'ios. }%
%
\en{For this reason, all widely used skill estimators are based on pairwise comparisons. }%
\es{Por esta raz\'on, todos los estimadores de habilidad ampliamente usados se basan en comparaciones por pares. }%
%
\en{Since the first generative models, proposed almost a century ago by~\cite{Thurstone1927} and~\cite{Zermelo1929}, it is assumed that the probability of the observed result $r$ depends on the performance $p$ of an agent $i$ and their opponent $j$, expressed as $P(\, r \,|\, p_i, \, p_j \,)$. }%
\es{Desde los primeros modelos generativos, propuestos hace casi un siglo por~\cite{Thurstone1927} y~\cite{Zermelo1929}, se supone que la probabilidad de un resultado observado $r$ depende del rendimiento $p$ del agente $i$ y de su oponente $j$, expresada como $P(\, r \,|\, p_i, \, p_j \,)$. }%
%
\en{The field continued to progress with the work of~\cite{Bradley1952} and~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}, leading to a breakthrough that took place when~\cite{Elo2008} developed a methodology for the US Chess Federation (USCF), which is still used by the International Chess Federation (FIDE). }%
\es{El campo sigui\'o progresando con los trabajos de \cite{Bradley1952} y~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}, que condujeron al gran avance que tuvo lugar cuando~\cite{Elo2008} desarrolló una metodología para la Federaci\'on de Ajedrez de los Estados Unidos (USCF), adoptada hasta el d\'ia de hoy por la Federaci\'on Internacional de Ajedrez (FIDE). }%

@@

\en{All currently used skill estimators share some variant of the probabilistic model proposed by Elo~\citep{Glickman1999,Herbrich2007,VanDerLinden2016,Fox2010}. }%
\es{Todos los estimadores de habilidad ampliamente utilizados en la actualidad comparten alguna variante del modelo probabil\'istico propuesto por Elo~\citep{Glickman1999, Herbrich2007, VanDerLinden2016, Fox2010}. }%
%
\begin{figure}[ht!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r = (d > 0)$};

@@

    \edge {d} {r};
    \edge {p1,p2} {d};
    \edge {s1} {p1};
    \edge {s2} {p2};
    %\node[invisible, right=of p2, xshift=4.35cm] (s-dist) {};
}
     \caption{
     \en{Generative model in which skills cause the observable results mediated by the difference of hidden performances, $d =p_i - p_j$, both random variables around their unknown true skill, $p \sim \N(s,\beta^2)$. }%
    \es{Modelo generativo en el que las habilidades causan los resultados observables a trav\'es de la diferencia de rendimientos ocultos, $d=p_i-p_j$, ambas variables aleatorias centradas en la verdadera habilidad, $p \sim \N(s,\beta^2)$. }%
    %
    \en{The one with the highest performance wins, $r = (d > 0)$. }%
    \es{Quien haya obtenido mayor rendimiento gana, $r = (d > 0)$. }%
    %
    \en{Observable variables are painted gray, hidden in white, and constants are shown as black dots. }%
    \es{Las variables observables se pintan de gris, las ocultas en blanco, y las constantes se muestran como puntos negros. }%
    }
    \label{fig:generative_model}
\end{figure}
%
\en{Figure~\ref{fig:generative_model} provides a causal model in which skills generates the observable result. }%
\es{La figura~\ref{fig:generative_model} ofrece un modelo causal en el que las habilidades generan el resultado observable. }%
%
\en{The agents exhibit different performances at each event, varying around their true skill, $\N(p\,|\,s,\beta^2)$. }%
\es{Los agentes exhiben distintos desempe\~nos en cada evento, que var\'ian alrededor de su verdadera habilidad, $\N(p\,|\,s,\beta^2)$. }%
%
\en{The model assumes that the agent with the highest performance wins, $r = (p_i > p_j)$. In other words, whoever obtains a difference of performance greater than 0 wins, $r = (p_i - p_j > 0)$. }%
\es{El modelo supone que gana el agente con mayor rendimiento, $r = (p_i > p_j)$. En otras palabras, gana quien obtenga una diferencia de desempeño mayor a 0, $r = (p_i - p_j > 0)$. }%
%
\en{The parameter $\beta^2$ is the same for all the agents and acts as the scale of the estimates: skill estimates which differ by $\beta$ are associated with the same probability of winning independently of the absolute value of the estimates. }%
\es{El parámetro $\beta^2$, al ser el mismo para todos los agentes, act\'ua como la escala de las estimaciones: habilidades separada por un $\beta$ mantienen siempre la misma probabilidad de ganar independiente del valor absoluto de las estimaciones. }%

@@

\en{Using the graphical representation proposed in Figure~\ref{fig:generative_model}, it is possible to derive the prediction of the result given the previous estimates, $P(\,r\,|\,s_{i_\text{old}},s_{j_\text{old}}\,)$. }%
\es{A partir de la representaci\'on gr\'afica propuesta en la figura \ref{fig:generative_model}, se puede derivar la predicci\'on del resultado dadas las estimaciones previas, $P(\,r\,|\,s_{i_\text{old}},s_{j_\text{old}}\,)$. }%
\en{Elo's methodological solution is simple and smart: updates both previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, based on the surprise, i.e. the complement of the prediction of the observed result. }%
\es{La soluci\'on metodol\'ogica de Elo es simple y astuta: actualizar las estimaciones previas, $s_{i_\text{old}}$ y $s_{j_\text{old}}$, en base a la sorpresa, i.e. el complemento de la predicci\'on del resultado observado. }%

@@

\en{where the probability arises from instantiating the previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, in the generative model (details in section~\ref{sec:2vs2}). }%
\es{Donde la probabilidad surge de instanciar las estimaciones previas en el modelo generativo (detalles en la secci\'on~\ref{sec:2vs2}). }%
\en{The idea is that the magnitude of the surprise $\Delta$ is related to the accuracy of the previous estimates and, therefore, can be used to update them. }%
\es{La idea es que la magnitud de la sorpresa $\Delta$ est\'a relacionada con cuan buenas son las estimaciones previas, y por lo tanto puede usarse para actualizarlas. }%
\en{Unexpected results would indicate that current estimates should be updated to a greater extent than if they had occurred as expected. }%
\es{Resultados inesperados indicar\'ian que las estimaciones actuales no son del todo correctas y deber\'ian actualizarse en mayor medida que si hubieran ocurrido como se esperaba. }%

@@

\en{where the surprise $\Delta$ acts as a correction factor for both previous estimates. }%
\es{Donde la sorpresa $\Delta$ act\'ua como factor de correcci\'on para ambas estimaciones previas. }%
\en{This procedure can recover the relative scale of the agents, starting from arbitrary initial values. }%
\es{Esta soluci\'on puede recuperar la escala relativa de los agentes, partiendo de valores iniciales arbitrarios. }%
\en{However, it does have a major weakness. }%
\es{Sin embargo, tiene algunas debilidades importantes. }%
\en{The update rule, shown in equation~\eqref{eq:elo_update}, is symmetric: what one agent wins is lost by the other. }%
\es{La regla de actualizaci\'on (Ec.~\eqref{eq:elo_update}) es sim\'etrica, as\'i que lo que gana un agente el otro lo pierde. }%
\en{Because new agents start with arbitrary skills (the same initial value is used for all individuals), they tend to generate high surprise in the initial events  and therefore can abruptly modify the opponents' estimates even though they have already converged. }%
\es{Debido a que a los agentes nuevos comienzan con estimaciones arbitrarias (el mismo valor inicial para cualquier individuo), ellas tienden a generan alta sorpresa y por lo tanto pueden modificar bruscamente las estimaciones de sus oponentes a pesar de que ya hubieran convergido. }%
\en{This weakness emerges because the uncertainties of the agent's estimates are not considered. }%
\es{Esta debilidad ocurre por no tener en cuenta la incertidumbre sobre las estimaciones de los agentes. }%
\en{An ad-hoc solution was proposed to solve this problem: reducing the impact of the surprise based on the number of times the agent has participated previously. }%
\es{Una soluci\'on ad-hoc fue propuesta para resolver este problema: reducir el impacto de la sorpresa en funci\'on de la cantidad de veces que el agente ha participado previamente. }%
% \en{This is the role played by the K-factor used by FIDE,  $\Delta_i = \Delta \cdot K_i$. }%
% \es{Ese es rol que desempe\~na el K-factor usado por la FIDE, $\Delta_i = \Delta \cdot K_i$. }%

@@

\en{A better alternative is to distribute our belief among all possible skill hypotheses, instead of selecting a single value as an estimate. }%
\es{Una alternativa superadora consiste en distribuir nuestra creencia entre todas las posibles hipótesis de habilidad, en vez de seleccionar un \'unico valor como estimaci\'on. }%
\en{This approach known as \emph{Bayesian inference} has proven successful in practice~\citep{Bishop2006} and ensures consistent reasoning in uncertainty contexts~\citep{Jaynes2003,VanHorn2003}. }%
\es{Este enfoque conocido como inferencia bayesiana ha demostrado ser exitoso en la pr\'actica~\citep{Bishop2006} y garantiza un razonamiento consistente en contextos de incertidumbre~\citep{Jaynes2003,VanHorn2003}. }%
\en{Any inference of this type, independently of its complexity, can be solved by the rules of probability: the~\ref{eq:sum_rule} and the~\ref{eq:product_rule}. }%
\es{Toda inferencia de este tipo, no importa cuan compleja sea, puede ser resuelta mediante las reglas de la probabilidad: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}. }%
\en{The \ref{eq:sum_rule} states that any marginal distribution can be obtained by integrating, or summing up, the joint distribution: }%
\es{La \ref{eq:sum_rule} afirma que cualquier distribuci\'on marginal se puede obtener integrando o sumando la distribuci\'on conjunta. }%

@@

\en{where $ P(\cdot)$ and $p(\cdot)$ represent discrete and continuous probabilities distributions respectively. }%
\es{Donde $P(\cdot)$ y $p(\cdot)$ representan distribuciones de probabilidad discretas y continuas respectivamente. }%
\en{Additionally, the \ref{eq:product_rule} states that any joint distribution can be expressed as the product of one-dimensional conditional distributions. }%
\es{Adem\'as, la \ref{eq:product_rule} se\~nala que cualquier distribuci\'on conjunta puede ser expresada como el producto de distribuciones condicionales uni-dimensionles. }%

@@

% \en{The inferential use of Bayes' theorem plays a central role in modern statistical learning techniques. }%
% \es{El uso inferencial del teorema de bayes juega un rol central en las t\'ecnicas modernas de aprendizaje estad\'isitico. }%
\en{The \ref{eq:bayes_theorem} optimally updates our beliefs about the hypotheses,  maximizing uncertainty after incorporating the information provided by the model and the data. }%
\es{El \ref{eq:bayes_theorem} actualiza de forma \'optima las creencia sobre las hip\'otesis, maximizando la incertidumbre luego de haber incorporar la información provista por el modelo y los datos. }%

@@

\en{where the only free variable is the skill hypothesis of agent $i$. }%
\es{Donde la \'unica variable libre es la hip\'otesis de habilidad del agente $i$. }%
\en{The prior quantifies the uncertainty about the skill before seeing the result, and the posterior quantifies the uncertainty after seeing the result. }%
\es{El prior cuantifica la incertidumbre sobre la habilidad antes de ver el resultado, y el posterior cuantifica la incertidumbre luego de ver el resultado. }%
